---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  # note: out.width must stay numeric-free; webshot2 receives it as vwidth
  # when screenshotting htmlwidgets and requires integer pixels
  screenshot.opts = list(vwidth = 1200, vheight = 700)
)
```

# dplyneage

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/tgerke/dplyneage/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tgerke/dplyneage/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

dplyneage draws interactive column-level lineage diagrams for dplyr and dbplyr pipelines. Pipe a query into `extract_lineage()` and it traces every output column back to the source columns it came from — through joins, aggregations, CTEs, unions, and computed expressions — then renders the result as a draggable, zoomable [React Flow](https://reactflow.dev/) diagram with `lineage_flow()`.

dbplyr pipelines are analyzed in pure R by walking their lazy query tree, so no Python is involved. Raw SQL goes through [sqlglot](https://github.com/tobymao/sqlglot)'s dedicated lineage engine instead, which means many dialects (DuckDB, PostgreSQL, Snowflake, BigQuery, ...) work too.

## Installation

```{r, eval = FALSE}
pak::pak("tgerke/dplyneage")
```

dbplyr pipelines need no Python at all. For raw SQL input, the Python dependency (sqlglot) is provisioned automatically the first time it's needed, via `reticulate::py_require()` — there is no setup step. See `vignette("python-integration")` if you manage your own Python environment.

## Usage

Build a dplyr pipeline against a database as usual, then pipe it into `extract_lineage()` and `lineage_flow()`:

```{r message=FALSE, fig.alt="Column-level lineage diagram with the customers and orders tables on the left and the summarised output table on the right, with edges tracing each output column back to its source columns"}
library(dplyneage)
library(dplyr)
library(dbplyr)
library(duckdb)

con <- dbConnect(duckdb::duckdb(), ":memory:")

customers <- tibble(
  customer_id = 1:5,
  first_name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  last_name = c("Smith", "Jones", "Brown", "Wilson", "Davis"),
  email = paste0(tolower(first_name), "@example.com")
)

orders <- tibble(
  order_id = 1:10,
  customer_id = rep(1:5, each = 2),
  amount = c(100, 150, 200, 75, 300, 125, 180, 90, 250, 160)
)

copy_to(con, customers, "customers", overwrite = TRUE)
copy_to(con, orders, "orders", overwrite = TRUE)

tbl(con, "customers") |>
  select(customer_id, first_name, last_name, email) |>
  left_join(tbl(con, "orders"), by = "customer_id") |>
  group_by(customer_id, first_name, last_name, email) |>
  summarise(
    total_orders = n(),
    total_spent = sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  extract_lineage() |>
  lineage_flow(height = "600px")
```

Behind that one pipe, `extract_lineage()`:

- walks the pipeline's lazy query tree in pure R, tracing every output column to its source columns (joins, aggregations, unions, and multi-source computed columns all resolve exactly)
- falls back to sqlglot's lineage engine when the pipeline injects raw SQL with `dbplyr::sql()`, or when you pass a SQL string directly (that path handles aliases, CTEs, and subqueries, and reads table schemas from your connection so unqualified columns attribute correctly)

The resulting diagram is fully interactive: drag tables to rearrange, zoom and pan, and hover columns to highlight their connections.

## Building diagrams by hand

For documentation or design work, you can construct lineage diagrams directly with `create_table_node()` and `create_column_edge()`:

```{r fig.alt="Hand-built lineage diagram showing the customers and orders source tables in blue connected to a customer_summary target table in green, with a SUM() label on the total_spent edge"}
nodes <- list(
  create_table_node(
    table_name = "customers",
    columns = c("customer_id", "name", "email"),
    x = 0, y = 50,
    table_type = "source"
  ),
  create_table_node(
    table_name = "orders",
    columns = c("order_id", "customer_id", "total_amount"),
    x = 0, y = 300,
    table_type = "source"
  ),
  create_table_node(
    table_name = "customer_summary",
    columns = c("customer_id", "customer_name", "total_spent"),
    x = 500, y = 150,
    table_type = "target"
  )
)

edges <- list(
  create_column_edge("customers", "customer_id", "customer_summary", "customer_id"),
  create_column_edge("customers", "name", "customer_summary", "customer_name"),
  create_column_edge("orders", "total_amount", "customer_summary", "total_spent",
                     label = "SUM()", animated = TRUE)
)

lineage_flow(nodes, edges, height = "600px")
```

Table types follow the color conventions used by dbt and SQLMesh:

| Type | Color | Use case |
|------|-------|----------|
| `source` | Blue | Raw/source tables |
| `transform` | Orange | Intermediate transformations |
| `target` | Green | Final output/materialized tables |

## Works with ducklake

Because `extract_lineage()` accepts any dbplyr lazy table, it composes directly with packages that produce them — for example [ducklake](https://github.com/tgerke/ducklake-r) tables:

```{r, eval = FALSE}
library(ducklake)

get_ducklake_table("orders") |>
  dplyr::left_join(get_ducklake_table("customers"), by = "customer_id") |>
  dplyr::group_by(customer_id) |>
  dplyr::summarise(total = sum(amount, na.rm = TRUE)) |>
  extract_lineage() |>
  lineage_flow()
```

The [ducklake lineage vignette](https://tgerke.github.io/dplyneage/articles/ducklake-lineage.html) works through a full example: building a small lake, diagramming each layer of a bronze/silver/gold pipeline, and extracting lineage from time-travel queries.

## Exporting lineage

Diagrams are for people; the same lineage is also useful as plain data. `lineage_json()` gives you a small, stable document you can query with jq, feed to a data catalog, or commit next to your pipeline code — then a CI diff flags any change to column provenance before it ships:

```{r}
lineage <- tbl(con, "orders") |>
  left_join(tbl(con, "customers"), by = "customer_id") |>
  group_by(customer_id, first_name) |>
  summarise(total_spent = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage()

lineage_json(lineage)
```

Written to a file, that document is scriptable from outside R entirely — here's jq answering "which source columns feed `total_spent`?":

```{r}
lineage_json(lineage, "lineage.json")
```

```{bash}
jq -r '.edges[] | select(.target_column == "total_spent")
       | "\(.source).\(.source_column)"' lineage.json
```

```{r, include = FALSE}
unlink("lineage.json")
```

`lineage_graphml()` writes GraphML, which opens directly in graph tools like Gephi, yEd, and igraph. The same question works as a graph query — and scales to transitive ancestry when pipelines chain:

```{r}
path <- tempfile(fileext = ".graphml")
lineage_graphml(lineage, path)

g <- igraph::read_graph(path, format = "graphml")
igraph::subcomponent(g, "output.total_spent", mode = "in")
```

## Learn more

- `vignette("getting-started")` walks from a first diagram through CTEs, multi-source columns, and schemas
- `vignette("python-integration")` covers how the Python dependency is managed
- Full function reference at [tgerke.github.io/dplyneage](https://tgerke.github.io/dplyneage/)
