---
title: "Column lineage for ducklake pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Column lineage for ducklake pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
can_eval <- requireNamespace("ducklake", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("dbplyr", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE)

# The time-travel example renders raw SQL, which needs the sqlglot engine
can_eval_sqlglot <- can_eval &&
  tryCatch(dplyneage::has_sqlglot(), error = function(e) FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = can_eval
)
```

[ducklake](https://github.com/tgerke/ducklake-r) gives you a versioned data lake: related tables stored together, every change snapshotted, full audit trails. dplyneage answers a question ducklake deliberately leaves to its query layer: *within a pipeline, where did each column come from?*

The two compose without any glue code. `get_ducklake_table()` returns an ordinary dbplyr lazy table, and that is exactly what `extract_lineage()` accepts — so any query you can write against a lake, you can also diagram.

## Setting up a small lake

We'll build a lake with two related tables. If you already have one, skip ahead and pipe your own queries.

```{r message=FALSE}
library(dplyneage)
library(ducklake)
library(dplyr)

install_ducklake()

lake_dir <- file.path(tempdir(), "lineage_lake")
dir.create(lake_dir, showWarnings = FALSE)
attach_ducklake("demo_lake", lake_path = lake_dir)
```

Load customers and their orders, with commit messages so the lake's audit trail says why each table exists:

```{r}
customers <- data.frame(
  customer_id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  region = c("east", "west", "east", "south", "west")
)

orders <- data.frame(
  order_id = 1:10,
  customer_id = rep(1:5, each = 2),
  amount = c(100, 150, 200, 75, 300, 125, 180, 90, 250, 160)
)

with_transaction(
  create_table(customers, "customers"),
  author = "Data Engineer",
  commit_message = "Load customer master"
)

with_transaction(
  create_table(orders, "orders"),
  author = "Data Engineer",
  commit_message = "Load order transactions"
)
```

## Lineage for a lake pipeline

Here is a typical analysis query — join orders to customers, aggregate by region — piped straight into lineage extraction:

```{r}
get_ducklake_table("orders") |>
  left_join(get_ducklake_table("customers"), by = "customer_id") |>
  group_by(region) |>
  summarise(total_sales = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage() |>
  lineage_flow(height = "400px")
```

Two things worth noticing:

- **Lake tables appear as blue source nodes under their plain names** (`orders`, `customers`) — the same names you passed to `create_table()`, with no catalog prefix cluttering the diagram.
- **No Python was involved.** Because ducklake queries are dbplyr pipelines under the hood, dplyneage's pure-R engine reads column provenance directly from the query tree. `total_sales` attributes to `orders.amount` and `region` to `customers.region`, exactly, with nothing to install beyond the two packages.

## Layered lakes: one diagram per hop

ducklake encourages a layered (bronze/silver/gold) architecture where each layer is materialized with `create_table()`. Lineage slots naturally into that workflow: the pipeline that *builds* a layer is also the recipe for its lineage diagram, so you can document the hop at the moment you materialize it.

```{r}
region_sales <- get_ducklake_table("orders") |>
  left_join(get_ducklake_table("customers"), by = "customer_id") |>
  group_by(region) |>
  summarise(total_sales = sum(amount, na.rm = TRUE), .groups = "drop")

with_transaction(
  create_table(region_sales, "region_sales"),
  author = "Data Analyst",
  commit_message = "Gold layer: sales by region"
)
```

A downstream query now sees `region_sales` as its source, and its lineage diagram documents just that hop:

```{r}
top_regions <- get_ducklake_table("region_sales") |>
  filter(total_sales > 300)

top_regions |>
  extract_lineage() |>
  lineage_flow(height = "300px")
```

## The whole lake in one diagram

Per-hop diagrams are useful at the moment you build a layer; a named list stitches them into the full picture. Pass each layer's build pipeline under the name it was materialized as, and `extract_lineage()` connects the hops: any source table matching a model name becomes an intermediate transform node.

```{r}
lake_lineage <- extract_lineage(list(
  region_sales = region_sales,
  top_regions = top_regions
))

lineage_flow(lake_lineage, height = "450px")
```

Now `orders` and `customers` (blue) feed `region_sales` (orange), which feeds `top_regions` (green) — the whole lake, bronze to gold, with every column edge preserved. Aggregation edges carry their defining expression as a label, so `total_sales` reads as `sum(amount, na.rm = TRUE)` right on the diagram.

The stitched graph is also where impact analysis pays off. `lineage_upstream()` walks the hops transitively:

```{r}
lineage_upstream(lake_lineage, "top_regions.total_sales")
```

So a rename of `orders.amount` touches every layer of this lake — worth knowing before you commit it.

## Time travel and lineage

ducklake's snapshots version your *data*; lineage describes your pipeline's *structure*. They compose: you can extract lineage from a time-travel query just like any other.

```{r, eval = can_eval_sqlglot}
first_version <- list_table_snapshots("orders")$snapshot_id[[1]]

get_ducklake_table_version("orders", first_version) |>
  group_by(customer_id) |>
  summarise(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage() |>
  lineage_flow(height = "300px")
```

One implementation detail surfaces here. `get_ducklake_table_version()` builds its query from raw SQL (`AT (VERSION => ...)` under the hood), which the pure-R engine can't see inside — so `extract_lineage()` falls back to the sqlglot engine, with a message. sqlglot parses DuckDB's time-travel syntax and traces the columns identically; the only practical difference is that this path uses dplyneage's Python dependency (provisioned automatically — see `vignette("python-integration")`).

The diagram is the same at every version, and that's the point: snapshots change what the data *was*, not where the columns *come from*. If a refactor ever does change provenance, `lineage_json()` gives you a stable document you can commit and diff in CI — see `vignette("getting-started")`.

```{r, include = FALSE}
detach_ducklake()
```

## Next steps

- ducklake's [time travel](https://tgerke.github.io/ducklake-r/articles/time-travel.html) and [transactions](https://tgerke.github.io/ducklake-r/articles/transactions.html) vignettes cover the lake side of this story
- `vignette("getting-started")` works through the cases where lineage gets hard: CTEs, multi-source columns, and `SELECT *`
- Found a lake query that traces incorrectly? Please [open an issue](https://github.com/tgerke/dplyneage/issues)
