---
title: "Getting Started with dplyneage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dplyneage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This guide walks you through setting up and using `dplyneage` to visualize column lineage in your dplyr/dbplyr pipelines.

## Installation

```{r, eval=FALSE}
pak::pak("tgerke/dplyneage")
```

## Setup

### Build React Flow Bundle (if needed)

The React Flow bundle should be pre-built and included with the package. However, if visualization doesn't work, you may need to build it yourself:

```{r, eval=FALSE}
library(dplyneage)

# Check if bundle exists
has_bundle()

# Build if needed (requires Node.js v18+)
build_bundle()
```

**For developers**: You can also build from the terminal in the package source directory:

```bash
./build_bundle.sh
```

### Install Python Dependencies

```{r, eval=FALSE}
install_sqlglot()
```

This will:

- Create a Python virtual environment
- Install sqlglot for SQL parsing and lineage extraction
- Configure the package to use this environment

**Important:** Restart R after installing Python dependencies.

### Verify Installation

```{r}
library(dplyneage)

# Check that sqlglot is available
has_sqlglot()
#> [1] TRUE

# Try the built-in example
lineage_example()
```

## Quick Start: Your First Lineage Visualization

### Simple Pipeline

```{r message=FALSE}
library(dplyr)
library(duckdb)

# Create a DuckDB connection
con <- dbConnect(duckdb(), ":memory:")

# Create sample data
customers <- tibble(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve")
)

orders <- tibble(
  order_id = 1:10,
  customer_id = rep(1:5, each = 2),
  amount = c(100, 150, 200, 75, 300, 125, 180, 90, 250, 160)
)

copy_to(con, customers, "customers", overwrite = TRUE)
copy_to(con, orders, "orders", overwrite = TRUE)

# Build a dplyr pipeline and visualize lineage
tbl(con, "customers") |>
  select(id, name) |>
  left_join(tbl(con, "orders"), by = c("id" = "customer_id")) |>
  group_by(id, name) |>
  summarise(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage() |>
  lineage_flow(height = "600px")
```

### Using Raw SQL

You can also extract lineage from raw SQL:

```{r}
sql <- "
SELECT 
  c.customer_id,
  c.name,
  SUM(o.amount) as total_spent
FROM customers c
LEFT JOIN orders o ON c.customer_id = o.customer_id
GROUP BY c.customer_id, c.name
"

extract_lineage(sql, dialect = "duckdb") |>
  lineage_flow()
```

### Manual Lineage Creation

For complete control, you can manually create lineage visualizations:

```{r}
# Define tables with their columns
source_table <- create_table_node(
  table_name = "customers",
  columns = c("id", "name", "email"),
  x = 0,
  y = 100,
  table_type = "source"
)

target_table <- create_table_node(
  table_name = "customer_summary",
  columns = c("customer_id", "full_name", "contact"),
  x = 500,
  y = 100,
  table_type = "target"
)

# Create column-to-column edges
edges <- list(
  create_column_edge("customers", "id", "customer_summary", "customer_id"),
  create_column_edge("customers", "name", "customer_summary", "full_name"),
  create_column_edge("customers", "email", "customer_summary", "contact")
)

# Visualize
lineage_flow(
  nodes = list(source_table, target_table),
  edges = edges
)
```

## Understanding the Workflow

The package follows this pipeline:

```
dplyr Pipeline (R code)
         ↓
dbplyr (SQL generation)
         ↓
sqlglot (Python parsing & column lineage)
         ↓
dplyneage (Graph creation)
         ↓
React Flow (Visualization)
```

## Interactive Features

Once your lineage is displayed, you can:

- **Drag nodes**: Click and drag tables to reposition them
- **Zoom**: Use mouse wheel to zoom in/out
- **Pan**: Click and drag the background to pan around
- **Select**: Click nodes or edges to select them

## Column Attribution

The package uses smart heuristics to determine which table each column comes from:

- Columns from `SELECT table.*` → attributed to `table`
- Standalone columns in JOINs → attributed to joined table
- Explicitly prefixed columns → attributed to specified table

Example:

```sql
SELECT customers.*, amount
FROM customers
LEFT JOIN orders ON (...)
```

Result:

- **customers**: `id`, `name` (from `customers.*`)
- **orders**: `amount` (standalone column from JOIN)
