---
title: "Getting Started with dplyneage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dplyneage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The raw-SQL examples need Python sqlglot, so only evaluate chunks when
# the full stack is available and the vignette builds anywhere
can_eval <- requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("dbplyr", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE) &&
  tryCatch(dplyneage::has_sqlglot(), error = function(e) FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = can_eval
)
```

dplyneage answers a simple question about your data pipelines: *where did each column come from?* You pipe a dplyr/dbplyr query (or pass raw SQL) into `extract_lineage()`, and render the answer as an interactive diagram with `lineage_flow()`.

This vignette starts with the smallest possible example and works up to the cases where lineage gets genuinely hard: joins with ambiguous columns, CTEs, and columns computed from several sources at once.

## Installation

```{r, eval = FALSE}
pak::pak("tgerke/dplyneage")
```

dplyr/dbplyr pipelines are analyzed entirely in R, so for those there is nothing else to install. Raw SQL strings are analyzed by sqlglot, dplyneage's one Python dependency — install the reticulate package to enable that engine, and sqlglot itself is provisioned automatically the first time it's needed. See `vignette("python-integration")` if you manage your own Python environment.

## Your first lineage diagram

Let's create a small in-memory DuckDB database to work with:

```{r message=FALSE}
library(dplyneage)
library(dplyr)
library(duckdb)

con <- dbConnect(duckdb(), ":memory:")

customers <- tibble(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  email = paste0(tolower(name), "@example.com")
)

orders <- tibble(
  order_id = 1:10,
  customer_id = rep(1:5, each = 2),
  amount = c(100, 150, 200, 75, 300, 125, 180, 90, 250, 160)
)

copy_to(con, customers, "customers", overwrite = TRUE)
copy_to(con, orders, "orders", overwrite = TRUE)
```

The simplest lineage there is — two columns selected from one table:

```{r}
tbl(con, "customers") |>
  select(id, name) |>
  extract_lineage() |>
  lineage_flow(height = "300px")
```

Each output column connects back to the source column it came from. Try dragging the tables around, zooming with the mouse wheel, and hovering a column to highlight its connections.

## A realistic pipeline

Lineage becomes useful once transformations pile up. Here is a join followed by an aggregation:

```{r}
tbl(con, "customers") |>
  left_join(tbl(con, "orders"), by = c("id" = "customer_id")) |>
  group_by(id, name) |>
  summarise(total_spent = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage() |>
  lineage_flow(height = "400px")
```

Notice that `total_spent` traces back to `orders.amount` — not to `customers`, even though `amount` appears unqualified in the generated SQL. When you pass a dbplyr table, dplyneage doesn't parse SQL at all: it walks the pipeline's own query tree, which records exactly which table each column came from, so attribution is always right.

If a pipeline embeds raw SQL with `dbplyr::sql()`, the query tree can't see inside that string, so `extract_lineage()` hands the whole query to sqlglot instead (with a message). You can also force a specific engine with `engine = "r"` or `engine = "sqlglot"` — see `?extract_lineage`.

## Where lineage gets hard

These are the cases that break naive lineage tools. dplyneage handles them because both engines resolve the full query structure rather than pattern-matching column names: dbplyr pipelines through their query tree, raw SQL through sqlglot's lineage engine. We'll use raw SQL here to keep the examples compact.

### Tracing through CTEs

Columns are traced *through* intermediate CTEs back to the base tables — `recent` is transparent, and `amount` correctly attributes to `orders`:

```{r}
extract_lineage("
  WITH recent AS (
    SELECT customer_id, amount FROM orders WHERE order_date > '2024-01-01'
  )
  SELECT customer_id, SUM(amount) AS total FROM recent GROUP BY customer_id
") |>
  lineage_flow(height = "300px")
```

### Columns with multiple sources

A computed column can come from several tables at once. `COALESCE` over a full join gets an edge from *both* sources:

```{r}
extract_lineage("
  SELECT COALESCE(u.email, a.email) AS email
  FROM users u FULL JOIN archive a ON u.id = a.id
") |>
  lineage_flow(height = "300px")
```

The same applies to arithmetic across tables (`o.amount * r.rate`), `CASE` expressions, and both branches of a `UNION`. dbplyr pipelines get the same treatment: a `full_join()` key column traces to both sides, and a `union_all()` column to every branch.

### Expanding `SELECT *`

With a schema available, `SELECT *` expands to real columns. dbplyr input never needs a schema — the pipeline itself knows its columns — but for raw SQL, pass one yourself:

```{r}
extract_lineage(
  "SELECT * FROM customers",
  schema = list(customers = c("id", "name", "email"))
) |>
  lineage_flow(height = "300px")
```

## Raw SQL and schemas

As the examples above show, `extract_lineage()` accepts a SQL string directly — useful for auditing queries you didn't write in R. Two things to know:

- **Qualified columns** (`o.amount`) always resolve correctly, schema or not.
- **Unqualified columns** need a schema to be attributed with certainty. Pass a named list mapping each table to its columns:

```{r}
extract_lineage(
  "SELECT c.name, order_date
   FROM customers c JOIN orders o ON c.id = o.customer_id",
  schema = list(
    customers = c("id", "name", "email"),
    orders = c("order_id", "customer_id", "order_date", "amount")
  )
) |>
  lineage_flow(height = "300px")
```

Without a schema, `SELECT *` cannot be expanded and produces a warning rather than a silently empty diagram.

Queries in other SQL dialects work by setting `dialect`:

```{r, eval = FALSE}
extract_lineage(query, dialect = "postgres")
extract_lineage(query, dialect = "snowflake")
```

## Building diagrams by hand

For documentation or design sketches, skip extraction entirely and build the diagram yourself:

```{r}
nodes <- list(
  create_table_node(
    table_name = "customers",
    columns = c("id", "name", "email"),
    x = 0, y = 100,
    table_type = "source"
  ),
  create_table_node(
    table_name = "customer_summary",
    columns = c("customer_id", "full_name", "contact"),
    x = 500, y = 100,
    table_type = "target"
  )
)

edges <- list(
  create_column_edge("customers", "id", "customer_summary", "customer_id"),
  create_column_edge("customers", "name", "customer_summary", "full_name"),
  create_column_edge("customers", "email", "customer_summary", "contact")
)

lineage_flow(nodes, edges, height = "300px")
```

Nodes come in three types — `"source"` (blue), `"transform"` (orange), and `"target"` (green) — and edges accept a `label` (e.g. `"SUM()"`) and `animated = TRUE` for emphasis. `lineage_example()` renders a complete hand-built diagram you can use as a template.

## Exporting lineage

Diagrams answer questions interactively; sometimes you need the same lineage as plain data. Two exporters cover the common cases.

`lineage_json()` serializes the nodes, edges, and metadata to a small, stable JSON document. Because the output is deterministic, you can commit it alongside your pipeline code and let CI diff it: if a refactor silently changes where a column comes from, the diff shows it before it ships. It is also the natural handoff format for data catalogs or anything scriptable with jq.

```{r}
lineage <- tbl(con, "customers") |>
  left_join(tbl(con, "orders"), by = c("id" = "customer_id")) |>
  group_by(id, name) |>
  summarise(total_spent = sum(amount, na.rm = TRUE), .groups = "drop") |>
  extract_lineage()

lineage_json(lineage)
```

`lineage_graphml()` writes GraphML, the XML format that graph tools speak: igraph, Gephi, and yEd all open it directly. Every column becomes its own node, which is what makes real graph queries possible. The classic one is impact analysis — "if `orders.amount` changes, which outputs are affected?" — or its reverse, tracing an output back to every source column that feeds it:

```{r, eval = can_eval && requireNamespace("igraph", quietly = TRUE)}
path <- tempfile(fileext = ".graphml")
lineage_graphml(lineage, path)

g <- igraph::read_graph(path, format = "graphml")

# Everything upstream of total_spent
igraph::subcomponent(g, "output.total_spent", mode = "in")

# Everything downstream of orders.amount
igraph::subcomponent(g, "orders.amount", mode = "out")
```

Both functions return the serialized string when called without `path`, so they compose in pipes and tests.

```{r, include = FALSE}
dbDisconnect(con)
```

## Next steps

- `vignette("python-integration")` explains how the Python dependency is managed, and how to use your own environment
- The [function reference](https://tgerke.github.io/dplyneage/reference/) documents every argument
- Found a query that traces incorrectly? Please [open an issue](https://github.com/tgerke/dplyneage/issues) with the SQL
