---
title: "Python Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The `dplyneage` package uses [sqlglot](https://github.com/tobymao/sqlglot) via the `reticulate` package to automatically extract column-level lineage from SQL queries. This follows the best practices outlined in the [reticulate package documentation](https://rstudio.github.io/reticulate/articles/package.html) for embedding Python dependencies in R packages.

## Installation

### First Time Setup

1. **Install the R package**:

```{r}
devtools::install()  # or load_all() for development
```

2. **Install Python dependencies**:

```{r}
library(dplyneage)
install_sqlglot()
```

The `install_sqlglot()` function will:

- Create a dedicated Python virtual environment (named `r-dplyneage` by default)
- Install `sqlglot` and its dependencies
- Configure the package to use this environment

3. **Restart R** to ensure the Python environment is properly loaded.

## Checking Installation

You can verify that sqlglot is available:

```{r}
library(dplyneage)

# Check if sqlglot is available
has_sqlglot()
#> [1] TRUE
```

If it returns `FALSE`, you may need to install or reconfigure:

```{r}
install_sqlglot()
```

## How It Works

### Architecture

```
R (dplyr/dbplyr) → SQL (via dbplyr::sql_render)
                        ↓
                 Python (sqlglot)
                        ↓
      Column Lineage Metadata (parse & analyze)
                        ↓
            R (create nodes & edges)
                        ↓
        React Flow Visualization
```

### Key Components

1. **R/zzz.R**: Package initialization and Python environment setup
   - `.onLoad()`: Initializes package-level variables
   - `configure_python_env()`: Lazy-loads sqlglot
   - `install_sqlglot()`: Installs Python dependencies
   - `has_sqlglot()`: Checks if sqlglot is available

2. **R/sqlglot_utils.R**: Core lineage extraction logic
   - `extract_lineage()`: Main user-facing function
   - `get_sql_from_dplyr()`: Converts dbplyr to SQL
   - `extract_lineage_from_sql()`: Calls Python sqlglot
   - `convert_lineage_to_graph()`: Creates visualization data

3. **inst/python/requirements.txt**: Python dependencies
   - Currently just `sqlglot>=23.0.0`

## Using Different Python Environments

By default, `dplyneage` creates its own virtual environment. You can customize this:

### Use a different environment name

```{r}
install_sqlglot(envname = "my-custom-env")
```

### Use conda instead of virtualenv

```{r}
install_sqlglot(method = "conda", envname = "r-dplyneage")
```

### Use an existing environment

If you already have sqlglot installed in a Python environment:

```{r}
# Point reticulate to your environment
Sys.setenv(RETICULATE_PYTHON = "/path/to/your/python")

# Then load the package
library(dplyneage)

# Check it works
has_sqlglot()
```

## SQL Dialects

sqlglot supports many SQL dialects. Specify the dialect when extracting lineage:

```{r}
# DuckDB (default)
extract_lineage(query, dialect = "duckdb")

# PostgreSQL
extract_lineage(query, dialect = "postgres")

# Snowflake
extract_lineage(query, dialect = "snowflake")

# BigQuery
extract_lineage(query, dialect = "bigquery")

# MySQL
extract_lineage(query, dialect = "mysql")
```

The dialect should match your database backend to ensure accurate parsing.

### Supported Dialects

Thanks to sqlglot, the package supports numerous SQL dialects:

- DuckDB (default - recommended for examples)
- PostgreSQL
- MySQL/MariaDB
- Snowflake
- BigQuery
- Redshift
- SQLite
- Oracle
- SQL Server
- And many more...

See the [sqlglot documentation](https://sqlglot.com/) for the full list.

## Performance

- **First call**: May be slow as Python environment initializes (~1-2 seconds)
- **Subsequent calls**: Fast (<100ms for typical queries)
- **Complex queries**: sqlglot can handle very complex SQL including:
  - CTEs (Common Table Expressions)
  - Subqueries
  - Window functions
  - Complex joins
  - Set operations (UNION, INTERSECT, etc.)

## Troubleshooting

### Python not found

If you get an error about Python not being found:

```{r}
# Install Miniconda (if needed)
reticulate::install_miniconda()

# Then install sqlglot
install_sqlglot()
```

### Import errors

If sqlglot imports fail:

```{r}
# Reinstall in a fresh environment
install_sqlglot(envname = "r-dplyneage-new")
```

### Check Python configuration

```{r}
# See which Python reticulate is using
reticulate::py_config()

# List installed packages
reticulate::py_list_packages()
```

## Development Notes

### Adding new Python dependencies

1. Add the package to `inst/python/requirements.txt`
2. Update `install_sqlglot()` if needed
3. Rebuild and reinstall the package
4. Run `install_sqlglot()` to update the environment

### Testing without installation

During development, you can test Python integration:

```{r}
devtools::load_all()

# This will warn if sqlglot isn't installed
has_sqlglot()

# Install if needed
install_sqlglot()
```

## References

- [Reticulate Package Documentation](https://rstudio.github.io/reticulate/)
- [Embedding Python in R Packages](https://rstudio.github.io/reticulate/articles/package.html)
- [sqlglot Documentation](https://sqlglot.com/)
- [sqlglot GitHub](https://github.com/tobymao/sqlglot)
