---
title: "Lineage that travels with the data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
can_eval <- requireNamespace("ducklake", quietly = TRUE) &&
  requireNamespace("pharmaversesdtm", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("dbplyr", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = can_eval
)
```

A [ducklake](https://github.com/tgerke/ducklake-r) snapshot records what a table's rows were at a point in time. It records nothing about how those rows were derived. That lives in code, in a git history the person reading the lake may not have, and the usual bridge is a git SHA in the commit message, which turns "how was this dataset built?" into "check out that commit and read it".

The two packages close the gap with no new mechanism on either side. A layer's column lineage is a small JSON document, `lineage_json()`. A ducklake commit accepts a free-text `commit_extra_info` field. Put the one in the other and the commit that materialized a table carries the derivation that produced it: `get_ducklake_table_version("adsl", v)` returns the rows as of snapshot `v`, and `lineage_from_json()` of the same snapshot returns the lineage, from the same catalog.

## When to do this

Someone has the lake and not the repository. A statistician checking a table, or an auditor reading a data cut, can ask where `adsl.TRT01A` came from without checking anything out, installing the pipeline, or opening a connection to anything but the lake.

A data cut and a code change look the same in the snapshot list. Both add a snapshot with a message. Lineage on the commit tells them apart: identical lineage means the rows moved and the derivation did not, and a changed lineage means the derivation moved, with `lineage_diff()` naming the columns.

The derivation is part of the record. For a regulated dataset, how `adsl` was built as of a cut is something a reviewer can ask about years later. Storing it on the commit versions it with the data, so snapshot retention applies to both together and the answer never depends on someone keeping the right commit of the right repository.

Two things this does not do. It records structure, not values: which rows changed between snapshots is ducklake's `get_table_changes()`, while lineage says which columns fed which. And it records what was written rather than gating it. `lineage_check()` on a pull request stops a breaking change before it lands (the [lineage checks in CI](https://tgerke.github.io/dplyneage/articles/lineage-ci.html) article); lineage on the commit documents the change after it did. A table written without the helper below carries no lineage, so the convention has to hold wherever the lake is written.

## Where lineage can live

The document `lineage_json()` writes has three homes, and each answers to a different reader.

| Where it lives | Who reads it | Article |
|---|---|---|
| The repository, as a committed file | `lineage_check()` on a pull request, before merge | [Lineage checks in CI](https://tgerke.github.io/dplyneage/articles/lineage-ci.html) |
| A catalog, as OpenLineage events | Marquez, DataHub, or any OpenLineage backend, beside dbt and Airflow lineage | [OpenLineage export and catalog round-trips](https://tgerke.github.io/dplyneage/articles/openlineage.html) |
| The lake, on the commit | Anyone with the lake, as of any snapshot | This article |

It is the same document each time, so a project can do all three from one extraction. The rest of this article is the third row.

## A lake and a publish helper

The lake lives in a temporary directory, and the article attaches it once. The data is two SDTM domains from [pharmaversesdtm](https://pharmaverse.github.io/pharmaversesdtm/): demographics and exposure.

```{r, message = FALSE}
library(dplyneage)
library(ducklake)
library(dplyr)

install_ducklake()

lake_dir <- file.path(tempdir(), "versioned_lake")
dir.create(lake_dir, showWarnings = FALSE)
attach_ducklake("versioned_lake", lake_path = lake_dir)
```

`publish()` materializes a layer and records its lineage on the same commit. It returns the snapshot id that commit produced.

```{r}
publish <- function(recipe, name, message) {
  lineage <- if (inherits(recipe, "tbl_lazy")) {
    lineage_json(extract_lineage(setNames(list(recipe), name)), pretty = FALSE)
  }
  with_transaction(
    replace_table(recipe, name),
    author = "pipeline",
    commit_message = message,
    commit_extra_info = lineage
  )
  invisible(max(list_table_snapshots()$snapshot_id))
}
```

Two choices in the helper matter. `pretty = FALSE` writes the document as a single line, which is the shape a metadata column wants. And the recipe reaches `extract_lineage()` as a one-element named list rather than on its own, so the output node takes the table's name: the stored lineage says `adsl.TRT01A` rather than `output.TRT01A`, and documents from several tables can be stitched together later.

Bronze layers are data frames, and a data frame has no query tree to read, so they carry no lineage; their commit message says where they came from. Demographics are complete from the first extract. Exposure records accrue as subjects are dosed, so that domain takes a cut date.

```{r, message = FALSE}
publish(pharmaversesdtm::dm, "dm", "Bronze: demographics")

ex_cut <- filter(pharmaversesdtm::ex, EXSTDTC <= "2013-12-31")
publish(ex_cut, "ex", "Bronze: exposure through 2013-12-31")
```

## The first cut

The gold layer keeps a few demographic columns, joins the first dose date from exposure, and derives planned and actual treatment, an age group, and a safety flag. The derivations are written as dplyr so the hop stays a query. admiral's versions work on data frames and leave no query tree behind; ducklake-r's [Clinical Trial Data Lake](https://tgerke.github.io/ducklake-r/articles/clinical-trial-datalake.html) article shows that side of the workflow.

```{r}
build_adsl <- function() {
  first_dose <- get_ducklake_table("ex") |>
    filter(!is.na(EXSTDTC)) |>
    group_by(USUBJID) |>
    summarise(TRTSDT = min(EXSTDTC, na.rm = TRUE), .groups = "drop")

  get_ducklake_table("dm") |>
    select(USUBJID, AGE, SEX, ARM, ACTARM) |>
    left_join(first_dose, by = "USUBJID") |>
    mutate(
      TRT01P = ARM,
      TRT01A = ACTARM,
      AGEGR1 = case_when(AGE < 65 ~ "<65", AGE < 80 ~ "65-79", TRUE ~ ">=80"),
      SAFFL = if_else(is.na(TRTSDT), "N", "Y")
    )
}
```

```{r, message = FALSE}
v1 <- publish(build_adsl(), "adsl", "Gold: ADSL, first cut")
```

The snapshot list shows the lineage riding along. Nothing here prints the document itself; `lineage_from_json()` is the way to read it.

```{r}
list_table_snapshots("adsl") |>
  transmute(snapshot_id, commit_message, lineage_chars = nchar(commit_extra_info))
```

## Reading it back

`lineage_at()` finds the snapshot and hands its document to `lineage_from_json()`. The check matters in a real lake: a snapshot written by anything other than `publish()` has no document, and the helper should say so rather than fail inside the parser.

```{r}
lineage_at <- function(table, version) {
  snapshots <- list_table_snapshots(table)
  doc <- snapshots$commit_extra_info[snapshots$snapshot_id == version]
  if (length(doc) != 1 || is.na(doc)) {
    stop("Snapshot ", version, " of ", table, " carries no lineage.")
  }
  lineage_from_json(doc)
}
```

No extraction ran to draw this diagram. It came from the catalog:

```{r}
lineage_flow(lineage_at("adsl", v1), height = "400px")
```

And this is the pairing the whole arrangement exists for: the rows and the derivation, both keyed by one snapshot id.

```{r}
get_ducklake_table_version("adsl", v1) |>
  count(SAFFL) |>
  collect()

lineage_upstream(lineage_at("adsl", v1), "adsl.SAFFL")
```

What comes back is the same object `extract_lineage()` would have returned, so everything that accepts a lineage accepts it: `lineage_flow()` for the diagram, `lineage_upstream()` and `lineage_downstream()` for impact questions, `lineage_diff()` and `lineage_check()` between two snapshots, and `lineage_openlineage()` to send one snapshot's lineage on to a catalog.

`get_ducklake_table_version()` builds its query from raw SQL. Tracing that query would send `extract_lineage()` to the sqlglot engine, as the [ducklake lineage](https://tgerke.github.io/dplyneage/articles/ducklake-lineage.html) article shows, but here it only fetches rows. The lineage was extracted from the recipe on the R engine when the layer was published, so this article runs without Python.

## A data refresh

Six months on, more subjects have been dosed. The exposure cut moves and the gold layer is rebuilt from the same recipe.

```{r, message = FALSE}
ex_cut <- filter(pharmaversesdtm::ex, EXSTDTC <= "2014-06-30")
publish(ex_cut, "ex", "Bronze: exposure through 2014-06-30")

v2 <- publish(build_adsl(), "adsl", "Gold: ADSL, second cut")
```

The safety flag moved with the data:

```{r}
get_ducklake_table_version("adsl", v2) |>
  count(SAFFL) |>
  collect()
```

The lineage did not:

```{r}
lineage_diff(lineage_at("adsl", v1), lineage_at("adsl", v2))
```

The snapshot list shows two commits with two messages. The lineage is what says the second one changed rows and nothing else.

## A refactor

Someone tidying `build_adsl()` points actual treatment at the planned arm. The pipeline still runs, the table is republished, and the twelve subjects whose actual arm differs from their randomized arm get a different `TRT01A`.

```{r, message = FALSE}
build_adsl <- function() {
  first_dose <- get_ducklake_table("ex") |>
    filter(!is.na(EXSTDTC)) |>
    group_by(USUBJID) |>
    summarise(TRTSDT = min(EXSTDTC, na.rm = TRUE), .groups = "drop")

  get_ducklake_table("dm") |>
    select(USUBJID, AGE, SEX, ARM, ACTARM) |>
    left_join(first_dose, by = "USUBJID") |>
    mutate(
      TRT01P = ARM,
      TRT01A = ARM,
      AGEGR1 = case_when(AGE < 65 ~ "<65", AGE < 80 ~ "65-79", TRUE ~ ">=80"),
      SAFFL = if_else(is.na(TRTSDT), "N", "Y")
    )
}

v3 <- publish(build_adsl(), "adsl", "Gold: ADSL, tidied derivations")
```

This time the diff is not empty:

```{r}
lineage_diff(lineage_at("adsl", v2), lineage_at("adsl", v3))
```

The removed edge is breaking because `adsl` is the table the lake's consumers read, and `TRT01A` is one of the columns they read from it. `lineage_check()` is the same comparison raised as an error. In CI it is a gate; on the lake it is a question asked after the fact, and the question "which commit changed the derivation?" has an answer from the catalog alone:

```{r}
derivation_history <- function(table) {
  snapshots <- list_table_snapshots(table)
  lineages <- lapply(snapshots$commit_extra_info, lineage_from_json)
  changed <- vapply(seq_along(lineages)[-1], function(i) {
    lineage_has_changes(lineage_diff(lineages[[i - 1]], lineages[[i]]))
  }, logical(1))
  data.frame(
    snapshot_id = snapshots$snapshot_id,
    commit_message = snapshots$commit_message,
    lineage_changed = c(NA, changed)
  )
}

derivation_history("adsl")
```

Three snapshots, one derivation change, and `list_table_snapshots()` on its own could not have said which.

## Before adopting it

The document here is a few kilobytes, and DuckLake stores `commit_extra_info` as unbounded text, so size is not a constraint at this scale. If a lake's conventions keep commit metadata short, or lineage is wanted for tables that something other than the pipeline writes, a versioned `lineage` table keyed by table name and snapshot id holds the same document; ducklake-r's clinical trial article stores define.xml that way. `lineage_from_json()` reads it back from either place.

Snapshot expiry removes the lineage with the snapshot. That is the same lifecycle as the data, which is the point: the derivation stays available for exactly as long as the rows it describes.

A [targets](https://tgerke.github.io/dplyneage/articles/targets-lineage.html) pipeline gets this with a one-line change. That article's `publish()` is this one without the `commit_extra_info` argument. Add it and every layer's commit carries its lineage, while `project_lineage()` there keeps reading the store as before.

```{r, include = FALSE}
detach_ducklake("versioned_lake", shutdown = TRUE)
```

## Next steps

- The [ducklake lineage](https://tgerke.github.io/dplyneage/articles/ducklake-lineage.html) article covers per-hop diagrams, stitching a whole lake, and lineage from time-travel queries
- The [targets](https://tgerke.github.io/dplyneage/articles/targets-lineage.html) article runs a bronze/silver/gold lake as a pipeline and separates data refreshes from provenance changes with `tar_outdated()` beside `lineage_diff()`
- The [lineage checks in CI](https://tgerke.github.io/dplyneage/articles/lineage-ci.html) article fails the pull request that would introduce a breaking change
- The [OpenLineage](https://tgerke.github.io/dplyneage/articles/openlineage.html) article sends the same document into a catalog
- ducklake's [time travel](https://tgerke.github.io/ducklake-r/articles/time-travel.html) and [transactions](https://tgerke.github.io/ducklake-r/articles/transactions.html) vignettes cover the lake side
- Found a snapshot whose lineage reads back wrong? Please [open an issue](https://github.com/tgerke/dplyneage/issues)
