---
title: "Project-level lineage for targets pipelines"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
can_eval <- requireNamespace("targets", quietly = TRUE) &&
  requireNamespace("pharmaversesdtm", quietly = TRUE) &&
  requireNamespace("ducklake", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("dbplyr", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = can_eval
)

# targets reads _targets.R from, and writes _targets/ under, the working
# directory, so the article gets a project directory of its own and every
# later chunk runs there. Start it empty: a store left behind by an earlier
# knit would make the first tar_outdated() come back with nothing to build.
proj_dir <- file.path(tempdir(), "targets_lineage")
unlink(proj_dir, recursive = TRUE)
dir.create(proj_dir)
knitr::opts_knit$set(root.dir = proj_dir)
```

dbt and SQLMesh compute lineage for a whole project. dplyneage works on what you hand it in a session: one query, or a named list of them. In R the project is a [targets](https://docs.ropensci.org/targets/) pipeline, and targets already knows the half of the story dplyneage doesn't: which targets are out of date and what a change will rerun. Its graph stops at the target, though. It can say that `adae` depends on `adsl`. It can't say which column of `adsl` feeds `TRTEMFL`.

The two compose without a new API. A target's value can carry the lazy query that built it, `tar_read()` hands that query back with no database attached, and `extract_lineage()` on a named list of those queries stitches the project into one diagram.

This article builds a small clinical-trial lake as a targets pipeline (three SDTM domains from [pharmaversesdtm](https://pharmaverse.github.io/pharmaversesdtm/), materialized as bronze, silver, and gold layers with [ducklake](https://github.com/tgerke/ducklake-r)) and runs it three times: an interim data cut, a later cut where only adverse events have accrued, and a code change. targets reports what reran. dplyneage reports whether the provenance moved. Those turn out to be different questions.

## A lake and a pipeline

The lake lives in the project directory, and the article attaches it once.

```{r, message = FALSE}
library(dplyneage)
library(targets)
library(ducklake)
library(dplyr)

install_ducklake()

lake_dir <- file.path(proj_dir, "lake")
dir.create(lake_dir)
attach_ducklake("trial_lake", lake_path = lake_dir)
```

Each target builds one layer of the lake and returns a small handle rather than the table itself:

- `table` is the name the layer was materialized under. Downstream targets read the lake by this name.
- `snapshot` is ducklake's commit id for that write. When new data lands, the id changes, targets sees a changed value, and everything downstream reruns.
- `recipe` is the lazy dplyr query that built the layer. Bronze layers are loaded from data frames and have none. Silver and gold layers do, and the recipe is what dplyneage reads.

The handle keeps the targets graph honest. `build_adae(ae, adsl)` names the two targets it depends on, reads their tables by name, and never composes on the upstream query, so each layer's lineage is one hop: the tables it read and the columns it made from them.

```{r}
publish <- function(recipe, name, message) {
  with_transaction(
    replace_table(recipe, name),
    author = "targets",
    commit_message = message
  )
  handle <- list(
    table = name,
    snapshot = max(list_table_snapshots()$snapshot_id)
  )
  if (inherits(recipe, "tbl_lazy")) {
    handle$recipe <- recipe
  }
  handle
}

read_layer <- function(handle) {
  get_ducklake_table(handle$table)
}

load_bronze <- function(data, name) {
  publish(data, name, paste("Bronze:", name))
}

clean_dm <- function(raw) {
  read_layer(raw) |>
    select(USUBJID, AGE, SEX, RACE, ARM, ACTARM) |>
    filter(ARM != "Screen Failure") |>
    mutate(across(c(SEX, RACE, ARM, ACTARM), ~ na_if(.x, ""))) |>
    publish("dm", "Silver: demographics")
}

clean_ex <- function(raw) {
  read_layer(raw) |>
    select(USUBJID, EXDOSE, EXSTDTC, EXENDTC) |>
    mutate(across(c(EXSTDTC, EXENDTC), ~ na_if(.x, ""))) |>
    publish("ex", "Silver: exposure")
}

clean_ae <- function(raw) {
  read_layer(raw) |>
    select(USUBJID, AESEQ, AEDECOD, AESEV, AESER, AESTDTC) |>
    mutate(across(c(AEDECOD, AESEV, AESTDTC), ~ na_if(.x, ""))) |>
    publish("ae", "Silver: adverse events")
}

build_adsl <- function(dm, ex) {
  first_dose <- read_layer(ex) |>
    filter(!is.na(EXSTDTC)) |>
    group_by(USUBJID) |>
    summarise(TRTSDT = min(EXSTDTC, na.rm = TRUE), .groups = "drop")

  read_layer(dm) |>
    left_join(first_dose, by = "USUBJID") |>
    mutate(
      TRT01P = ARM,
      TRT01A = ACTARM,
      AGEGR1 = case_when(AGE < 65 ~ "<65", AGE < 80 ~ "65-79", TRUE ~ ">=80"),
      SAFFL = if_else(is.na(TRTSDT), "N", "Y")
    ) |>
    publish("adsl", "Gold: subject-level analysis dataset")
}

build_adae <- function(ae, adsl) {
  read_layer(ae) |>
    inner_join(
      read_layer(adsl) |> select(USUBJID, TRT01A, TRTSDT, SAFFL),
      by = "USUBJID"
    ) |>
    mutate(TRTEMFL = if_else(AESTDTC >= TRTSDT, "Y", "N")) |>
    publish("adae", "Gold: adverse-event analysis dataset")
}
```

The silver step does the blank-to-`NA` cleaning that admiral's `convert_blanks_to_na()` would do, written as dplyr so the hop stays a query. The gold derivations are deliberately plain: `ADSL` gets planned and actual treatment, an age group, first dose date, and a safety flag; `ADAE` joins the treatment start onto each event and flags the treatment-emergent ones. Real derivations go through admiral, whose functions work on data frames and so leave no query tree behind; ducklake-r's [Clinical Trial Data Lake](https://tgerke.github.io/ducklake-r/articles/clinical-trial-datalake.html) article shows that side of the workflow.

## The interim cut

Demographics and exposure are locked from the first extract. Adverse events keep accruing through follow-up, so the pipeline takes a cut date for that domain and nothing else. The date is an ordinary R object that the `ae_raw` command refers to, and targets hashes it like any other global.

```{r}
ae_cut_date <- "2013-12-31"

tar_script({
  tar_option_set(packages = c("dplyr", "ducklake"))
  list(
    tar_target(dm_raw, load_bronze(pharmaversesdtm::dm, "dm_raw")),
    tar_target(ex_raw, load_bronze(pharmaversesdtm::ex, "ex_raw")),
    tar_target(
      ae_raw,
      load_bronze(filter(pharmaversesdtm::ae, AESTDTC <= ae_cut_date), "ae_raw")
    ),
    tar_target(dm, clean_dm(dm_raw)),
    tar_target(ex, clean_ex(ex_raw)),
    tar_target(ae, clean_ae(ae_raw)),
    tar_target(adsl, build_adsl(dm, ex)),
    tar_target(adae, build_adae(ae, adsl))
  )
}, ask = FALSE)

tar_outdated(callr_function = NULL, reporter = "silent")
```

One choice needs explaining. targets normally runs `_targets.R` in a fresh process, which is what makes a pipeline reproducible. This article runs it in the current session instead (`callr_function = NULL`), for two reasons: the layer functions above are defined here rather than in an `R/` directory, and a ducklake catalog is a single DuckDB file that two processes cannot both hold open for writing. The last section covers what changes in a real project.

```{r, message = FALSE}
tar_make(callr_function = NULL, reporter = "silent")
tar_progress()
```

## What a target holds

Bronze handles carry a name and a snapshot. Gold handles carry the query too.

```{r}
str(tar_read(dm_raw))

names(tar_read(adsl))
```

The recipe came back from the store, and its connection did not. `tar_make()` saved it with `saveRDS()`, and a DuckDB connection does not survive that round trip, so printing the recipe would fail when dplyr tried to run a preview. `show_query()` and `extract_lineage()` only read the query tree, which is why lineage from a store needs no database at all.

```{r}
adsl_recipe <- tar_read(adsl)$recipe

show_query(adsl_recipe)
```

On its own, a recipe diagrams a single hop, the same as any lazy table:

```{r}
adsl_recipe |>
  extract_lineage() |>
  lineage_flow(height = "400px")
```

## The whole project in one diagram

The project graph is the recipes of every completed target, named by the table each one materialized, passed to `extract_lineage()` as a list. That naming is what stitches the hops together: `adae` reads a table called `adsl`, and a model in the list is called `adsl`, so the two connect.

```{r}
project_lineage <- function() {
  handles <- lapply(sort(tar_meta(targets_only = TRUE)$name), tar_read_raw)
  handles <- Filter(function(h) !is.null(h$recipe), handles)
  recipes <- lapply(handles, function(h) h$recipe)
  names(recipes) <- vapply(handles, function(h) h$table, character(1))
  extract_lineage(recipes)
}

interim <- project_lineage()
lineage_flow(interim, height = "600px")
```

Bronze handles have no recipe and never enter the list, but the silver recipes read their tables, so `dm_raw`, `ex_raw`, and `ae_raw` appear as blue source nodes. `adsl` is orange because `adae` reads it. `adae` is green because nothing does.

Impact analysis works across the hops. A treatment-emergent flag depends on dates from two domains, three layers back:

```{r}
lineage_upstream(interim, "adae.TRTEMFL")
```

## A later data cut

Six months on, the adverse-event extract is longer and nothing else has changed. Moving the cut date is the only edit.

```{r}
ae_cut_date <- "2014-06-30"

tar_outdated(callr_function = NULL, reporter = "silent")
```

targets traced the new value to `ae_raw` and from there to the two targets that read it. The other five are skipped:

```{r, message = FALSE}
tar_make(callr_function = NULL, reporter = "silent")
tar_progress()
```

The lake now has a second snapshot of `ae_raw`, `ae`, and `adae`. The lineage is the same graph it was, because rows changed and columns did not:

```{r}
refreshed <- project_lineage()
lineage_has_changes(lineage_diff(interim, refreshed))
```

`tar_outdated()` and `lineage_diff()` look at the same run and answer different questions. One says the data moved. The other says the structure didn't. For a study database that gets a new cut every few weeks, that second answer is the one a reviewer wants to see, and it costs nothing to produce.

## A code change

Now a refactor. Someone tidying `build_adsl()` points actual treatment at the planned arm, a one-word change that alters numbers for the twelve subjects whose actual arm differs from their randomized arm. The pipeline still runs.

```{r}
build_adsl <- function(dm, ex) {
  first_dose <- read_layer(ex) |>
    filter(!is.na(EXSTDTC)) |>
    group_by(USUBJID) |>
    summarise(TRTSDT = min(EXSTDTC, na.rm = TRUE), .groups = "drop")

  read_layer(dm) |>
    left_join(first_dose, by = "USUBJID") |>
    mutate(
      TRT01P = ARM,
      TRT01A = ARM,
      AGEGR1 = case_when(AGE < 65 ~ "<65", AGE < 80 ~ "65-79", TRUE ~ ">=80"),
      SAFFL = if_else(is.na(TRTSDT), "N", "Y")
    ) |>
    publish("adsl", "Gold: subject-level analysis dataset")
}

tar_outdated(callr_function = NULL, reporter = "silent")
```

targets hashes functions as well as objects, so the edit invalidates `adsl` and, through it, `adae`. The bronze and silver layers stay put:

```{r, message = FALSE}
tar_make(callr_function = NULL, reporter = "silent")
tar_progress()
```

This time the diff is not empty:

```{r}
refactored <- project_lineage()
lineage_diff(refreshed, refactored)
```

The removed edge is breaking because `adsl.TRT01A` feeds `adae.TRT01A`, and a column that something downstream consumes has changed its source. `lineage_check()` turns that into an error:

```{r, error = TRUE}
lineage_check(refreshed, refactored, annotate = FALSE)
```

We pass `annotate = FALSE` because this article renders on a GitHub Actions runner, where the default prints workflow commands instead of plain lines. In a pull request that failure is the point: the [lineage checks in CI](https://tgerke.github.io/dplyneage/articles/lineage-ci.html) article has the job that runs it.

## Taking this to a real project

Three things move when this leaves the article. The layer functions go into `R/` and `_targets.R` loads them with `tar_source()`. The lake is attached inside `_targets.R`, so the pipeline process is the only one holding the catalog during a build. And `tar_make()` runs with its default fresh process.

```r
# _targets.R
library(targets)
tar_option_set(packages = c("dplyr", "ducklake"))
tar_source()
attach_ducklake("trial_lake", lake_path = "lake")
ae_cut_date <- "2014-06-30"

list(
  tar_target(dm_raw, load_bronze(pharmaversesdtm::dm, "dm_raw")),
  tar_target(ex_raw, load_bronze(pharmaversesdtm::ex, "ex_raw")),
  tar_target(ae_raw, load_bronze(read_ae_extract(ae_cut_date), "ae_raw")),
  tar_target(dm, clean_dm(dm_raw)),
  tar_target(ex, clean_ex(ex_raw)),
  tar_target(ae, clean_ae(ae_raw)),
  tar_target(adsl, build_adsl(dm, ex)),
  tar_target(adae, build_adae(ae, adsl))
)
```

`project_lineage()` stays as it is. It reads only the store, so it can run after `tar_make()` in the same CI job with the lake already closed, and its result is the value the lineage check compares between branches.

```{r, include = FALSE}
detach_ducklake("trial_lake", shutdown = TRUE)
tar_option_reset()
```

## Next steps

- The [ducklake lineage](https://tgerke.github.io/dplyneage/articles/ducklake-lineage.html) article covers per-hop diagrams and lineage from time-travel queries
- ducklake-r's [Clinical Trial Data Lake](https://tgerke.github.io/ducklake-r/articles/clinical-trial-datalake.html) article builds the full SDTM-to-ADaM lake, admiral derivations included
- The [targets manual](https://books.ropensci.org/targets/) covers `tar_source()`, cues, and everything else about running the pipeline
- Found a pipeline that stitches wrong? Please [open an issue](https://github.com/tgerke/dplyneage/issues)
